Scaffolder - Software for Manual Genome Scaffolding.

### 2011-09-13 src/article.tex
Update author details
Further proof read

\author{
  Michael D Barton$^{1,2}$\correspondingauthor%
  \email{Michael D Barton\correspondingauthor  - mail@michaelbarton.me.uk}%
\and
  Hazel A Barton$^{1,2}$%
  \email{Hazel A Barton - bartonh@nku.edu}%
      }

\address{
\iid(1) Biology Department, The University of Akron, Akron, %
OH 44325-3908, USA \\
\iid(2) Department of Biological Sciences, Northern Kentucky %
University, Nunn Drive, Highland Heights, KY 41076, USA}

@@

  \paragraph*{Background:} The assembly of next-generation short-read
  sequencing data can result in a fragmented non-contiguous set of genomic
  sequences. Therefore a common step in a genome project is to join
  neighbouring sequence regions together and fill gaps . This scaffolding step
  is non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining these sequences together also hides the source of each
  region in the final genome sequence. Taken together these considerations
  may make reproducing or editing an existing genome scaffold difficult.

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
or biased and incomplete sequencing data, may prevent the genome being
assembled into a continuous sequence. This may be due to insufficient, or
multiple different overlaps between reads and is analogous to missing pieces
in the jigsaw or pieces that fit to multiple other pieces. \pb

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence. This sequence is useful for representing both the known
and to-be-resolved areas in the genome sequence.

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs, combined with additional gap filling sequences
produced by PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in YAML specify how these sequences should be joined.
The Scaffolder software is used to generate the scaffold sequence from these
instructions. In addition to specifying which contigs are required, the
scaffold file allows the contigs to be edited into smaller sub-sequences or
reverse complemented if necessary. Each scaffold file represents one scaffolded
nucleotide sequence and separate scaffolds should be defined in separate files.
\pb

@@

The nucleotide sequences used in the scaffold are maintained as a separate
FASTA file. The nucleotide sequences are referenced in the scaffold using the
first word from the FASTA header of the corresponding sequence. Maintaining the
nucleotide sequences in a separate file preserves the unedited sequence and
decouples the data from the specification of how it should be used to produce
the genome sequence. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, indicating a region of unknown sequence but known length.
The second line specifies the size of this unknown region. This entry produces
a region of 20 `N' characters in the scaffold. \pb

@@

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either the `open' or `close' tags is required when
adding an insert sequence. If only one of the `open' or `close' tags is used
the corresponding opposing `open'/`close' coordinate is calculated from the
length of the insert FASTA sequence. This allows inserts to bridge into, and
partially fill gap regions, without requiring an end coordinate position. \pb

@@

Constructing a genome by specifying the scaffold organisation in text file
makes generating a scaffold super sequence both reproducible and deterministic
for the same file and set of FASTA sequences. In comparison, joining large
nucleotide sequences by hand cannot be reliably reproduced. The scaffold file
furthermore provides a human readable description of how the scaffold is
constructed. Configuring the final sequence in the scaffold file means the
build is easier to edit once already constructed. \pb

A use case for Scaffolder is a combination of computational and manual editing
of a genome scaffold. We have used Scaffolder in our own genome projects to
create an initial scaffold from computationally parsing the output of \emph{in
silico} scaffolding tools into YAML. This scaffold was then manually updated as
the scaffold was finished with additional gap-filling sequences generated in
the laboratory. This is example of the scaffolder format being both
computationally tractable while being simple to edit manually. \pb

@@

Scaffolder is software aimed at both bioinformaticians and biologists familiar
with the command line who wish to build a genome scaffold from a set of
contigs. The Scaffolder file format maintains the genome scaffold as a concise
and readable text representation that allows third-parties to see how the
genome sequence was scaffolded. This file format also allows a broad overview
of which sequences were included and how they are ordered in the genome
scaffold, something not possible to deduce from a megabase-length string of
nucleotide characters. Scaffolder furthers increases the ease of
reproducibility in genome projects by allowing the scaffold super-sequence to
be reliably reproduced from the same scaffold file. The YAML syntax for writing
the scaffold file is also standardised and simple to manipulate
programmatically. This thereby means the scaffolding process follows the Unix
tenet of ``If your data structures are good enough, the algorithm to manipulate
them should be trivial.''

>>>>

\author{
  Michael D Barton$^{1}$\correspondingauthor%
  \email{Michael D Barton\correspondingauthor  - mail@michaelbarton.me.uk}%
\and
  Hazel A Barton$^{1}$%
  \email{Hazel A Barton - bartonh@uakron.edu}%
      }

\address{
\iid(1) Biology Department, The University of Akron, Akron, %
OH 44325-3908, USA}

@@

  \paragraph*{Background:} The assembly of next-generation short-read
  sequencing data can result in a fragmented non-contiguous set of genomic
  sequences. Therefore a common step in a genome project is to join
  neighbouring sequence regions together and fill gaps. This scaffolding step
  is non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining these sequences together also hides the source of each
  region in the final genome sequence. Taken together these considerations
  may make reproducing or editing an existing genome scaffold difficult.

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
or biased and incomplete sequencing data may prevent the genome being
assembled into a continuous sequence. This may be due to insufficient or
multiple different overlaps between reads and is analogous to missing pieces
in the jigsaw or pieces that fit to multiple other pieces. \pb

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters `N' allows a draft
continuous sequence. This sequence is useful for representing both the known
and to-be-resolved areas in the genome sequence.

@@

The input data for Scaffolder are nucleotide sequences in FASTA format file.
These nucleotide sequences can be of any length and may be individual reads,
assembled contigs or contigs which have been joined into larger scaffolds. The
case in which Scaffolder may be most useful is using the contigs and scaffolded
contigs, combined with additional gap filling sequences produced by PCR or
\emph{in silico} methods as outlined in the Background. \pb

@@

The Scaffolder software facilitates reproducibly joining nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in YAML specify how these sequences should be joined.
The Scaffolder software is used to generate the scaffold sequence from these
instructions. In addition to specifying which contigs are required, the
scaffold file allows the contigs to be edited into smaller sub-sequences or
reverse complemented if necessary. Each scaffold file represents one scaffolded
nucleotide sequence and as such separate scaffolds should be defined in
separate files. \pb

@@

The nucleotide sequences used in the scaffold are maintained as a separate
FASTA file: the nucleotide sequences are referenced in the scaffold using the
first word from the FASTA header of the corresponding sequence. Maintaining the
nucleotide sequences in a separate file preserves the unedited sequence and
decouples the data from the specification of how it should be used to produce
the genome sequence. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, indicating a region of unknown sequence but known length.
The second line specifies the size of this unknown region. In this example this
entry produces a region of 20 `N' characters in the scaffold. \pb

@@

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either the `open' or `close' tags is required when
adding an insert sequence. If only one of the `open' or `close' tags is used
the corresponding opposing `open'/`close' coordinate is calculated from the
length of the insert FASTA sequence. This allows inserts to bridge into, and
partially fill, gap regions without requiring an end coordinate position. \pb

@@

Constructing a genome by specifying the scaffold organisation in text file
makes generating a scaffold super sequence both reproducible and deterministic
for the same file and set of FASTA sequences. In comparison, joining large
nucleotide sequences by hand cannot be reliably reproduced, while the scaffold
file also provides a human readable description of how the scaffold is
constructed. Configuring the final sequence in the scaffold file means the
build is easier to edit, once constructed. \pb

An example use case for Scaffolder is a combination of computational and manual
editing of a genome scaffold. We have used Scaffolder in our own genome
projects to create an initial scaffold from computationally parsing the output
of \emph{in silico} scaffolding tools into YAML. This scaffold was then
manually updated as the scaffold was finished with additional gap-filling
sequences generated in the laboratory. This is example of the scaffolder format
being both computationally tractable while being simple to edit manually. \pb

@@

Scaffolder is software, written in Ruby, aimed at both bioinformaticians and
biologists familiar with the command line who wish to build a genome scaffold
from a set of nucleotide sequences. The Scaffolder file format maintains the
genome scaffold as a concise and readable text representation that allows
third-parties to see how the genome sequence was scaffolded. This file format
also allows a broad overview of which sequences were included and how they are
ordered in the genome scaffold, something not possible to deduce from a
megabase-length string of nucleotide characters. Scaffolder furthers increases
the ease of reproducibility in genome projects by allowing the scaffold
super-sequence to be reliably reproduced from the same scaffold file. The YAML
syntax for writing the scaffold file is also standardised and simple to
manipulate programmatically. This thereby means the scaffolding process follows
the Unix tenet of ``If your data structures are good enough, the algorithm to
manipulate them should be trivial.''
